---
title: "project1_jawlik"
author: "Matthew A Jawlik"
date: "6/23/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Loading and preprocessing the data

No initial pre-processing to do; will restructure as needed to address subsequent sections.

```{r, echo=T, warning=F, message=FALSE}
library(ggplot2)
library(mice)
library(VIM)

activity <- read.csv("activity.csv",header = T)
```



## What is mean total number of steps taken per day?

Here, we will aggregate the dataset by date--summing up the steps--and then perform calculations.

```{r, echo=TRUE, warning=F}
days <- aggregate(steps~date, data = activity, FUN = sum)
hist(days$steps, breaks = 20)
mean(days$steps)
median(days$steps)
```



## What is the average daily activity pattern?

Next, we will aggregate the origianl dataset now by interval--averageing the numbers steps--and then perform calculations.

```{r, echo=TRUE, warning=F}
ints <- aggregate(steps~interval, data = activity, FUN = mean)
qplot(interval, steps, data = ints, geom = "line")
```


The interval with the maximum number of steps is:
```{r, echo=TRUE, warning=F}
ints[ints$steps == max(ints$steps),]
```



## Imputing missing values

There are a lot NAs on the steps variable:

```{r, echo=TRUE, warning=F}
md.pattern(activity) # show the missing values
activity_plot <- aggr(activity, col=c('green','gray'),
                  numbers=TRUE, sortVars=TRUE,
                  labels=names(activity), cex.axis=.7,
                  gap=3, ylab=c("Missing data","Pattern"))
```


To address this we will use the mice package to create multiple (5) imputed datasets and then combine them into a master set.  The method we will use is "Predictive Mean Matching" (PMM), which has the benefit of working with non-normal variables.

```{r, echo=TRUE, warning=F}
activity_imputed <- mice(activity, m=5, maxit = 5, method = 'pmm')
activity_complete <- complete(activity_imputed)
```


Performing the same calculations as we did on the original dataset, we see the averages (means and medians) are now substantially higher.  Without imputing, our data would have under-reported these figures.  Imputing increased them to the correct actual levels.

```{r, echo=TRUE, warning=F}
days_complete <- aggregate(steps~date, data = activity_complete, FUN = sum)
hist(days_complete$steps, breaks = 20)
mean(days_complete$steps)
median(days_complete$steps)
```



## Are there differences in activity patterns between weekdays and weekends?

To answer these, we first determine the day of the week for each date, and then if that day is a weekday or weekend.  We then aggregate (average) based on that determination and plot.

```{r, echo=TRUE, warning=F}

activity_complete$date <- as.Date(activity_complete$date)
activity_complete$weekday <- weekdays(activity_complete$date)
activity_complete$dayType <- ifelse(activity_complete$weekday %in% c("Saturday", "Sunday"),
                                    "Weekend", "Weekday")
ints_complete <- aggregate(steps~interval+dayType, data = activity_complete, FUN = mean)
qplot(interval, steps, data = ints_complete, facets = dayType~., geom = "line")
```





